openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2021-03-25 paper/policar2020-opentsne.tex
Apply BMC template

% \documentclass[twocolumn]{article}
\documentclass[letter]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}

%%% Load packages
%\usepackage{amsthm}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

\usepackage{graphicx}
\usepackage{minted}
\usemintedstyle{vs}

\usepackage{amsmath,amssymb}
\usepackage{booktabs}

\newcommand{\opentsne}{\textsf{openTSNE}}

\renewcommand{\baselinestretch}{1.2}

@@

@@

\author{
  Pavlin G. Poli\v{c}ar\textsuperscript{1}\\
  \texttt{pavlin.policar@fri.uni-lj.si}\\
  \and
  Martin Stra\v{zar}\textsuperscript{2}\\
  \texttt{martin.strazar@gmail.com}\\
  \and
  Bla\v{z} Zupan\textsuperscript{1,3}\\
  \texttt{blaz.zupan@fri.uni-lj.si}\\
}

\date{
  \textsuperscript{1}Faculty of Computer and Information Science, University of Ljubljana,\\
  Ve\v{c}na pot 113, SI-1000 Ljubljana, Slovenia\\ \vspace{2mm}
  \textsuperscript{2}Broad Institute of Harvard and MIT,\\
  D\"{u}sternbrooker Weg 20, MA-02142 Cambridge, U.S.A.\\ \vspace{2mm}
  \textsuperscript{3}Department of Molecular and Human Genetics, Baylor College of Medicine,\\
  1 Baylor Plaza, TX-77030 Houston, U.S.A.
}

\maketitle

\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology are often used to reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t\nobreakdash-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t\nobreakdash-SNE algorithm and its many
extensions.
The library is fast and enables users to embed data sets containing millions of
data points in a matter of minutes on consumer-grade laptop computers.
Unique to \opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

\providecommand{\keywords}[1]{\vspace{2mm}\textbf{Keywords ---} #1}
\begin{keywords}
t-SNE, embedding, visualization, dimensionality reduction, Python library
\end{keywords}

\section*{Introduction}

@@

\section*{Conclusion}

@@

\section*{Availability}

\opentsne\ is distributed under the BSD-3-Clause License and is publicly
available as an open-source package at
\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
in this study are included in accompanying notebooks, publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

\section*{Competing interests}
The authors declare no competing interests.

\section*{Author's contributions}
P.G.P. implemented and maintains \opentsne\ library. M.S. helped in the development of mathematical foundations and advised on the inclusion of the extensions. B.Z. initiated and supervised the project. All three co-authors wrote the manuscript.

\section*{Acknowledgements}
This work was supported by Slovenian Research Agency (P2-0209).
We want to thank Dmitry Kobak for various helpful discussions and best practices
when using t-SNE and his contributions to the source code. We would also
like to thank George Linderman for his help with the FIt-SNE algorithm.

\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{references}      % Bibliography file (usually '*.bib' )

\eject
\fontsize{8}{10}\selectfont
\section*{Online Methods}

@@

>>>>

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

\documentclass[twocolumn]{bmcart}

\usepackage{amsthm,amsmath}
\usepackage[utf8]{inputenc}
\usepackage{url}

\usepackage{minted}
\usemintedstyle{vs}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\newcommand{\opentsne}{\textsf{openTSNE}}
\endlocaldefs

@@

\begin{frontmatter}
\begin{fmbox}
\dochead{Software}

@@

\author[
  addressref={aff1},
  corref={aff1},
  email={pavlin.policar@fri.uni-lj.si}
]{\inits{P.G.}\fnm{Pavlin G.} \snm{Poli\v{c}ar}}
\author[
  addressref={aff2},
  email={martin.strazar@gmail.com}
]{\inits{M.S.}\fnm{Martin} \snm{Stra\v{zar}}}
\author[
  addressref={aff1,aff3},
  email={martin.strazar@gmail.com}
]{\inits{B.Z.}\fnm{Bla\v{z}} \snm{Zupan}}

\address[id=aff1]{
  \orgdiv{Faculty of Computer and Information Science},
  \orgname{University of Ljubljana},
  \street{Ve\v{c}na pot 113},
  \city{Ljubljana},
  \cny{Slovenia}
}
\address[id=aff2]{%
  \orgname{Broad Institute of Harvard and MIT},
  \street{D\"{u}sternbrooker Weg 20},
  %\postcode{}
  \city{Cambridge},
  \cny{U.S.A.}
}
\address[id=aff3]{%
  \orgdiv{Department of Molecular and Human Genetics},
  \orgname{Baylor College of Medicine},
  \street{1 Baylor Plaza},
  %\postcode{}
  \city{Houston},
  \cny{U.S.A.}
}

%\end{fmbox}% comment this for two column layout

\begin{abstractbox}
\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology are often used to reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t\nobreakdash-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t\nobreakdash-SNE algorithm and its many
extensions.
The library is fast and enables users to embed data sets containing millions of
data points in a matter of minutes on consumer-grade laptop computers.
Unique to \opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

\begin{keyword}
\kwd{t-SNE}
\kwd{embedding}
\kwd{visualization}
\kwd{dimensionality reduction}
\kwd{Python library}
\end{keyword}

\end{abstractbox}

\end{fmbox}% uncomment this for two column layout

\end{frontmatter}

\section*{Background}

@@

\section*{Conclusions}

@@

%\fontsize{8}{10}\selectfont
\section*{Methods}

@@

\begin{backmatter}

\section*{Acknowledgements}
We want to thank Dmitry Kobak for various helpful discussions and best practices
when using t-SNE and his contributions to the source code. We would also
like to thank George Linderman for his help with the FIt-SNE algorithm.

\section*{Funding}
This work was supported by Slovenian Research Agency (P2-0209).

\section*{Availability of data and materials}
\opentsne\ is distributed under the BSD-3-Clause License and is publicly
available as an open-source package at
\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
in this study are included in accompanying notebooks, publicly available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Authors' contributions}
P.G.P. implemented and maintains \opentsne\ library. M.S. helped in the development of mathematical foundations and advised on the inclusion of the extensions. B.Z. initiated and supervised the project. All three co-authors wrote the manuscript.

\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{references}

\end{backmatter}
